---
title: "Visualizing FEMA NRI Data"
subtitle: "EDS 240 HW #2" 
author: Isabella Segarra 
date: "1/28/26"
format: html
---
# HW #2: Visualizing FEMA NRI Data

## Objective
In this assignment, we are tasked with answering the following question with our data visualization: 

*How do FEMA National Risk Index scores for counties in California compare to those in other states?*


```{r}
#| label: Set-Up
#| output: false
#| message: false 
#| warning: false 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(here)
library(ggplot2)
library(janitor)
library(ARTofR)
library(ggtext)

#..........................import data...........................
 nri_df <- read_csv(here::here("data", "National_Risk_Index_Counties_807384124455672111.csv")) %>% 
   clean_names()
 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            wrangle nri data                         
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

 nri_2 <- nri_df %>%
   # Filter for only US states
   filter(!state_name %in% c(
       "Puerto Rico",
       "Guam",
       "Virgin Islands",
       "Northern Mariana Islands",
       "American Samoa",
       "District of Columbia"
     )
   ) %>%
   # Group by state
   group_by(state_name) %>%
   # Count the total number of counties
   summarize(
     total_counties = n(),
     # Sum the counties considered high risk (percentile > 84.32)
     high_risk_counties = sum(
       national_risk_index_state_percentile_composite > 84.32,
       na.rm = TRUE # remove Nas
     ),
     # Find the proportion of high risk in each county
     prop_high_risk_counties = high_risk_counties / total_counties
   ) %>%
   # Create boolean column for data visualizing later 
   mutate(ca_true_false = state_name == "California")
 

 
```
## Build visualization

```{r}
#| fig-alt: "Lollipop graph of National Risk Index data depicting the top 10 states with a high proportion of counties at risk according to FEMA."
#| warning: false 

ca_prop_plot <- nri_2 %>%
  # Get the top 10 high risk states 
  slice_max(prop_high_risk_counties, n = 10) %>% 
  # Pull out prop_risk just for California
  # Plot proportion high risk 
  ggplot(aes(x = prop_high_risk_counties, 
             # reorder y-axis (state_name) based on prop high risk (highest go first)
             y = reorder(state_name, prop_high_risk_counties),
             # spotlight California 
             color = ca_true_false)) +
  # Lolipop graph = line + point 
  geom_linerange(aes(xmin = 0, xmax = prop_high_risk_counties), size = 1.1) + 
  geom_point(size = 3.5) +
  annotate("text", x = 0.79, y = "California", label = "0.79", hjust = -0.3, size = 4)+
  # Adjust colors for California ("TRUE") and non-California states ("FALSE")
  scale_color_manual(values = c("FALSE" = "#4A7BA7", "TRUE" = "#de6021")) +
  # Add labels 
  labs(title = "Proportion of High-Risk Counties: California vs. Top States", 
     subtitle = "California ranks 3rd with 79% of counties classified as high-risk",
     x = "Proportion of High Risk Counties", 
     caption = "High-risk defined as FEMA Risk Index percentile &gt; 84.32<br>*Data: FEMA National Risk Index (2025 Release)*") +
  theme_minimal() +
  # Theme adjustments 
 theme(axis.title.y = element_blank(), # Remove y label 
        legend.title = element_blank(), 
        legend.position = "none", # Remove legend 
        plot.title = element_text(hjust = 0.5, size = 15, lwd = 1.5, face = "bold"), 
       plot.subtitle = element_text(hjust = 0.5, size = 10),
       plot.caption = element_markdown(margin = margin(t = 15, unit = "pt")
       )
)

ca_prop_plot

# Save plot 
ggsave("figures/final_plot.png")


```

## Questions

**1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?**

My variables of interest were the `total_counties` (categorical), `national_risk_index_state_percentile_composite` (numeric), and `state_name`(categorical). I was interested in the `national_risk_index_state_percentile_composite` because of the background reading about FEMA's National Risk Index (NRI) where it was mentioned that a percentile of 84.32 means that the county has higher risk than 84.32% of counties nationwide. I wrangled the data to produce a variable called `prop_high_risk_counties` which is the proportion of counties within the state that are high risk. 


**2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?**

I originally began with a bar chart to represent this data, however I chose a lollipop graph because it draws your eyes to the `geom_point` where you can read the proportion values clearly. I chose to represent the data because it reduces the visual clutter and summarizes the main findings clearly while showing the ranking clearly. I decided to highlight the actual proportion of California counties at risk to further highlight that California is the focus of this plot. 


**3. Summarize your main finding in no more than two sentences.**
This graph shows that among the top 10 states, California ranks third with 79% of the state's counties classified as high-risk by FEMA's National Risk Index.


**4. What modifications did you make to this visualization to make it more easily readable?**
In order to make the visualization more readable, I added a subtitle to help interpret the proportion values. In addition, I decided to highlight the California lollipop line to show that the central state is California in this analysis. I decided to create a true/false column instead of using gghighlight because I wanted to keep the other state's lines prominent for this visualization. 


**5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.**
I wanted to change the color of California in the state y axes and perhaps in the title. 

