---
title: "Visualizing ACS Data"
subtitle: "EDS 240 HW #3" 
author: Isabella Segarra 
date: "2/12/26"
format: html
---
# HW #3: Visualizing FEMA NRI x ACS Data

## Objective 
In this assignment, we are tasked with answering the following question with our data visualization: 

*How does climate hazard risk exposure vary across racial / ethnic groups in California?*

I accomplished this task with two visualizations 1) Alluvial diagram and 2) Dodge Bar Chart. 

```{r}
#| label: Set-Up
#| output: false
#| message: false 
#| warning: false 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(here)
library(ggplot2)
library(janitor)
library(ARTofR)
library(ggtext)
library(tidycensus)
library(janitor)
library(ggalluvial)
library(showtext)
#..........................import data...........................

# ________Import ACS data________

#....Step 1a: see all available ACS variables + descriptions.....
#acs_vars <- tidycensus::load_variables(year = 2023, dataset = "acs1")

#..............Step 1b: import race & ethnicity data.............
# race_ethnicity <- tidycensus::get_acs(
#   geography = "county",
#   survey = "acs1",
#   # Define variables I may use 
#   variables = c("B01003_001", "B02001_002", "B02001_003", 
#                 "B02001_004", "B02001_005", "B02001_006",
#                 "B02001_007", "B02001_008", "B03002_012",
#                 "B03002_002"),
#   state = "CA", 
#   year = 2023) |>
#   # join variable descriptions 
#   dplyr::left_join(acs_vars, by = dplyr::join_by(variable == name)) 
# 
# #.................Step 2: write ACS data to file.................
# readr::write_csv(race_ethnicity, here::here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#..................Step 3: read in your CSV file.................
race_ethnicity <- readr::read_csv(here::here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#________Import FEMA data_________
nri_df <- read_csv(here::here("data", "National_Risk_Index_Counties_807384124455672111.csv")) %>% 
   clean_names()

#.....Import fonts......
font_add_google("Hanken Grotesk", "hanken")

```

## Data Wrangling
cleaning tasks:
- `str_remove()` the name column (remove California; new column 
- new column called `county_name`
- create new column `county_pop` w/ case_when (if label == 'estimate!total')
- create new column `race_ethnic` w/ case_when 
  -remove "estimatetotal!" from column 
- resulting `race_ethnicity` column should contain the following additional columns:  `county_name`, `county_pop`, `race_ethnic`

```{r}
#..........................Wrangle ACS Data ...........................
race_ethnicity_clean <- race_ethnicity %>% 
  # all columns in lower snack case
  clean_names() %>% 
  # remove "County,California" from column because all data is from California
   mutate(county_name = str_remove(name, " County, California"), 
          # Create total population of county column
          tot_pop_county = if_else(label == "Estimate!!Total", estimate, NA),
          # Create race_ethnicity column 
          race_ethnicity = if_else(
      label != "Estimate!!Total",
      str_remove(label, "Estimate!!Total:!!"),
      NA
    )) %>% 
  # Select only the columns you want 
  select(county_name, race_ethnicity, tot_pop_county, estimate) 

#..........................Wrangle NRI Data  ...........................

nri_3 <- nri_df %>%
   # Filter for California
   filter(state_name == "California") %>%
  # Select columns interested in
  select(county_name, population_2020, national_risk_index_score_composite, national_risk_index_rating_composite)

 
#..........................Join Data  ...........................  
nri_race <- full_join(race_ethnicity_clean, nri_3, by = "county_name")

#..........................Data prep..........................
nri_race_sankey <- nri_race %>% 
  # Reorder risk levels from low to high
  mutate(national_risk_index_rating_composite = fct_relevel(
    national_risk_index_rating_composite, 
    "Very Low",
    "Relatively Low", 
    "Relatively Moderate", 
    "Relatively High", 
    "Very High"
  )) %>%
  # Sum population across all counties for each race/ethnicity and risk level
  group_by(race_ethnicity, national_risk_index_rating_composite) %>%
  summarize(pop_risk_category = sum(estimate, na.rm = TRUE),
            .groups = "drop") %>% # Ungroup within summarize 
  # Calculate percentage of each racial/ethnic group in each risk category
  group_by(race_ethnicity) %>%
  mutate(
    total_pop = sum(pop_risk_category),
    pct_in_risk_category = pop_risk_category / total_pop * 100
    ) %>% 
  # Rename race/ethnicity column factors for simplified graphing 
  mutate(race_ethnicity = case_when(
    race_ethnicity == "American Indian and Alaska Native alone" ~ "Native American",
    race_ethnicity == "Black or African American alone" ~ "Black", 
    race_ethnicity == "Asian alone" ~ "Asian",
    race_ethnicity == "White alone" ~ "White", 
    race_ethnicity == "Native Hawaiian and Other Pacific Islander alone" ~ "Native Hawaiian/Pacific Islander",
    race_ethnicity == "Some Other Race alone" ~ "Other",
    race_ethnicity == "Two or More Races:" ~ "Mixed",
    race_ethnicity == "Not Hispanic or Latino:" ~ "Non-Hispanic",
    race_ethnicity == "Hispanic or Latino:" ~ "Hispanic/Latino"))

```

## Build Visualization
### Alluvial Diagram 
```{r}
#| fig-alt: "Alluvial diagram of National Risk Index and US Census data depicting climate risk exposure for certain ethnic and racial groups across California."
#| warning: false 
#..........................Palette..........................

risk_color_pallette <- c(
  "Relatively Low" = "#1BB6AFFF",      
  "Relatively Moderate" = "#fee090", 
  "Relatively High" = "#f46d43",     
  "Very High" = "#d73027"            
)

#..........................Graph..........................

nri_race_sankey %>% 
  # Filter out NAs
  filter(!is.na(race_ethnicity), !is.na(national_risk_index_rating_composite)) %>% 
  # Reorder by population 
  mutate(race_ethnicity = fct_reorder(race_ethnicity, total_pop, .desc = TRUE)) %>%
  # Plot 
  ggplot(aes(axis1 = race_ethnicity, 
           axis2 = national_risk_index_rating_composite,
           y = pop_risk_category)) +
  geom_alluvium(aes(fill = national_risk_index_rating_composite), 
                alpha = 0.8, 
                curve_type = "cubic") + # Sankey type 
  scale_fill_manual(values = risk_color_pallette) + # Custom color pallette defined above
  geom_stratum(width = 1/3) + # Width of each stratum 
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 2.5, 
            fontface = "bold", 
            check_overlap = TRUE) +  # Make sure labels don't overlap 
  # Define box titles and width 
  scale_x_discrete(limits = c("Race/Ethnicity", "Risk Level"), expand = c(0.12, 0.1)) +
  # Turn y axes into values wiht commas 
  scale_y_continuous(labels = scales::comma) +
  labs(fill = "FEMA Risk Index",
       y = "Population",
       x = NULL, # Remove x axis title
       title = "Climate Risk Exposure by Race and Ethnicity in California",
       subtitle = "Majority of Groups Face Severe Climate Risk Exposure", 
       caption = "Data: FEMA National Risk Index (2023 Release)\nand US Census Bureau Community Survey (ACS)") +
  # Theme adjustments 
  theme_minimal() +
  theme(aspect.ratio = 1, 
        plot.title = element_text(font = "hanken", hjust = 0.5, size = 15, lwd = 1.5, face = "bold"), 
       plot.subtitle = element_text(hjust = 0.5, size = 10), 
        plot.caption = element_text(hjust = 1)) 


```
### Dodge Bar Chart

```{r}
#| fig-alt: "Dodge bar chart of National Risk Index and US Census data depicting climate risk exposure for certain ethnic and racial groups across California. Data reveals that majority of groups face severe climate risk exposure in California."
#| warning: false 

# reorder columns by "Very High" risk 
race_order <- nri_race_sankey %>%
  filter(!is.na(race_ethnicity), national_risk_index_rating_composite == "Very High") %>%
  arrange(pct_in_risk_category) %>%
  pull(race_ethnicity) %>%
  as.character()

dodge_bar_chart <- nri_race_sankey %>%
  # Filter out NAs
  filter(!is.na(race_ethnicity), !is.na(national_risk_index_rating_composite)) %>%
  # Plot 
  ggplot(aes(x = factor(race_ethnicity, levels = race_order), y = pct_in_risk_category,
             fill = national_risk_index_rating_composite)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = risk_color_pallette) +
  # Flip x and y 
  coord_flip() +
  labs(fill = "FEMA Risk Index",
       y = "% Population in Risk Category",
       x = "Race/ethnicity",
       title = "Climate Risk Exposure by Race and Ethnicity in California",
       subtitle = "Majority of groups face severe climate risk exposure", 
       caption = "Data: FEMA National Risk Index (2023 Release)\nand US Census Bureau Community Survey (ACS)") +
  # Theme adjustments 
    theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_text(family = "hanken"),
        axis.text.x = element_text(family = "hanken"),
        axis.text.y = element_text(family = "hanken"),
        legend.title = element_text(family = "hanken"),
        plot.title = element_text(family = "hanken", hjust = 0.5, size = 15, lwd = 1.5, face = "bold", margin = margin(b = 10)), 
       plot.subtitle = element_text(family = "hanken", hjust = 0.5, size = 10), 
       plot.caption = element_text(family = "hanken", margin = margin(t = 15)), # Adjust margin spacing
       panel.grid.minor = element_blank()) 

# Enable show text
showtext_auto(enable = TRUE)

# View plot
dodge_bar_chart

```
```{r}
#..............enable {showtext} for newly opened GD.............
showtext_auto(enable = TRUE)

#...................set resolution to match GD...................
showtext_opts(dpi = 300)

#..............write plot to file (aka save as png)..............
ggsave(
  filename = here::here("figures", "hw3_plot.png"),
  plot = dodge_bar_chart,
  device = "png",
  width = 8, 
  height = 7,
  unit = "in",
  dpi = 300
)

#...............turn off {showtext} text rendering...............
showtext_auto(enable = FALSE)
```

## Questions
**1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?**

My variables of interest were `race_ethnicity` (categorical), `pop_risk_category` which is the population of each ethnic or racial group within `national_risk_index_composite` (categorical). I also created `pct_in_risk_category` which was the percentage of that population in the risk (numeric). 

**2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?**

I chose to represent this data with a bar chart because it most clearly shows the trends within demographics. Due to the high proportion of each racial and ethnic group being high risk, I chose to use dodge bar chart to show the large difference between *high risk* and *moderately risk* in each group. 

**3. Summarize your main finding in no more than two sentences.**

All demographic groups across California face high risk to climate hazards, particularly Black, Asian, and other race/ethnic groups. 

**4. What modifications did you make to this visualization to make it more easily readable?**
Modifications made to make this graph more interpretable was using `coord_flip` for the busy x axes labels, and used position = dodge to show the large difference between high risk and moderatley risk. 

**5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.**
For my alluvial diagram, I wanted to change the overlapping of text. I tried check_overlap = TRUE within the `geom_text()` argument, however that did not work. I also tried to rewrite the columns so that they do not overlap with `case_when()`, however this issue persists because the size of axis1 stratum  (population) is small for native american and pacific islander. I tried working with other packages, including `{alluvial}`, `{ggsankey}`, and `{networkD3}`. The last two are considered sankey diagrams, which differs from the alluvial diagram. Although I tried all the other packages, I prefer the `{ggalluvial}` package because it is an extension of ggplot2 and you are able to make alterations in the same way you do any ggplot. I would love to keep work shopping this graph! 
